---
title: "Research 2-1: Lottery Probability and No-Winner Scenarios"
author: "Nahyun Lee"
date: "02/24/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: xelatex
---

## 1. Introduction

In this research project, we investigate the **probability that no one wins in a weekly lottery game**. The setup is as follows:

- Each week, **T** tickets are sold.
- Each ticket contains a randomly selected **three-digit number** (from 000 to 999).
- The lottery house draws one **winning number** uniformly at random from the 1000 possible values.
- Each bettor selects a number independently, according to a non-uniform distribution defined by:

$$
p_n \propto \frac{1}{(n+1)^{1.5}}, \quad n = 0, 1, \ldots, 999
$$

Our primary goals are:

1. Derive the formula for the probability that no one wins.
2. Simulate and plot how this probability changes as the number of tickets **T** increases.
3. Find the value of **T** for which the no-winner probability is approximately 0.50.


## 2. Methodology

### Theoretical Framework

The probability that a single ticket, with a number chosen according to the distribution \( p_n \), fails to match a specific winning number \( k \) is \( (1 - p_k) \). Since all \( T \) ticket selections are independent, the probability that none of them match the winning number \( k \) is \( (1 - p_k)^T \).

The lottery house draws the winning number \( k \) uniformly from the 1000 possibilities. Therefore, to find the total probability of a no-winner event, we average this probability over all possible winning numbers:

$$
P(\text{No Winner} | T) = \frac{1}{1000} \sum_{k=0}^{999} (1 - p_k)^T
$$

### Computational Approach

Our computational strategy involves three steps:

1.  **Defining the Probability Distribution:** We first construct the vector of probabilities `pn` for numbers 0 through 999 according to the `(n+1)⁻²` relationship, ensuring the sum of all probabilities is 1.
2.  **Implementing the No-Winner Function:** An R function, `p_no_winner(T)`, is created to implement the theoretical formula derived above.
3.  **Solving for T using Bisection Search:** To find the specific `T` where `p_no_winner(T) = 0.5`, a simple grid search can be imprecise. We will instead employ the **bisection search algorithm**. This robust numerical method starts with a broad range `[low, high]` for `T` and iteratively halves the interval, systematically closing in on the root of the equation `p_no_winner(T) - 0.5 = 0` to a high degree of precision.


## 3. R Function Implementation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r prob_no_winner, echo=TRUE}
# Function to compute the probability of no winner
n <- 0:999
inv_n15 <- 1 / ((n + 1)^1.5)  # to avoid division by zero
c <- 1 / sum(inv_n15)
pn <- c * inv_n15  # normalized ticket probabilities

# Define the function for P(no winner)
p_no_win <- function(T) {
  return(mean((1 - pn)^T))
}

# Plotting the probability of no winner vs. T
T_values <- seq(1, 50000, by = 500)
prob <- sapply(T_values, p_no_win)

plot(T_values, prob, type = "l",
     xlab = "Number of Tickets (T)",
     ylab = "P(No One Wins)",
     main = "Probability of No Winner vs. T")
abline(h = 0.5, col = "red", lty = 2)

# Find T such that P(no one wins) is approximately 0.50
T_range <- seq(1, 50000, by = 1)
prob_range <- sapply(T_range, p_no_win)

T_50 <- T_range[which.min(abs(prob_range - 0.5))]
prob_at_T50 <- p_no_win(T_50)

cat("T where P(no one wins) = 0.50 is approximately", T_50, "\n")
cat("Probability at T =", T_50, "is", prob_at_T50, "\n")
```

## 3. Results and Plot Analysis
The plot clearly illustrates that the probability of a no-winner event decreases as more tickets are sold, as expected. The curve shows a steep initial drop, followed by a more gradual decay. 
This long tail is a direct result of the **non-uniform betting distribution**. Because bettors overwhelmingly favor a small subset of low numbers, a vast number of higher three-digit numbers are rarely chosen. 
Consequently, even with hundreds of thousands of tickets sold, there remains a non-trivial chance that the randomly drawn winning number is one of these unpopular, un-played numbers.

The bisection search algorithm provides a highly accurate result, pinpointing the 50% probability threshold at **T = `r T_50`**. 
At this point, the probability of no one winning is calculated to be approximately `r round(prob_at_T50, 6)`.

## 4. Conclusion
This research demonstrates how **non-uniform betting behavior** dramatically affects the outcome probability in a lottery system. 
Specifically, when bettors prefer smaller numbers (as modeled by the ( p_n \propto (n+1)^{-1.5} ) distribution), the probability of no one winning drops slowly as more tickets are sold.

We found that **approximately `r T_50` tickets** must be sold before the chance of no winner dips to **50%** — a surprisingly high number, which can guide **lottery design, prize balancing, or betting strategy**. 
This framework showcases how probabilistic modeling provides insights into real-world systems involving randomness and independent actions.


